---
title: "Discovery Module Report"
subtitle: "Sample **`r params$sample_name`**"
author: "MSc. Matheus Cosentino DeepVir v.1.0"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    smooth_scroll: true
    theme: yeti
params:
  sample_name: "DefaultSample" 
  fasta_file: "path/to/fasta"
  diamond_file: "path/to/diamond"
  darkmatter_file: "path/to/darkmatter"
  logos: "path/to/logo/" 
  output_dir: "path/to/output"
---

🔍This report aims to provide a detailed view of the Library **`r params$sample_name`** contig assembly, comparing the overall sequence quality with the success of **taxonomic annotation via Diamond** against the nr5 NCBI Database & iddentification on putative Viral agents that compose the **Viral Dark Matter** within the non-iddentifiaded contigs.

The analysis is structured into four parts: 

1. Overall **contig quality metrics**

2. Global taxonomic diversity per **Kingdom**

3. A hierarchical breakdown for **Viral Families** identified 

4. Report of Contigs within HMMs hits for the **RNA-dependent RNA Polymerase (RdRp)** 

Reproducibility of analysis and report are ensured by the DeepVir pipeline (https://github.com/matheus-cosentino/deepvir.git).

```{r functions-local, message=FALSE, warning=FALSE, echo = FALSE}
summarize_dnabin_lengths <- function(dnabin_object, object_label) {
  if (!is.list(dnabin_object) || length(dnabin_object) == 0) {
    return(data.frame(
      Metric = c(
        "N Sequences", "Total Bases (Mb)", "Mean Len.", "Median Len.",
        "Std. Dev.", "Min Len.", "Max Len.", "Q1 (25%)", "Q3 (75%)"
      ),
      Value = NA,
      row.names = NULL
    ) %>%
      rename(!!object_label := Value))
  }

  sequence_lengths <- sapply(dnabin_object, length)
  summary_stats <- summary(sequence_lengths)
  stdev <- sd(sequence_lengths)
  quantiles <- quantile(sequence_lengths, probs = c(0.25, 0.50, 0.75))

  # Define os formatos desejados (são todos strings, exceto o float)
  format_int <- "%d"
  format_float <- "%.3f"
  format_mb <- "%.2f"

  summary_df <- data.frame(
    Metric = c(
      "N Sequences", "Total Bases (Mb)", "Mean Len.", "Median Len.",
      "Std. Dev.", "Min Len.", "Max Len.", "Q1 (25%)", "Q3 (75%)"
    ),
    Value = c(
      # Valores que devem ser inteiros: usamos format_int
      sprintf(format_int, length(sequence_lengths)),
      # Total Bases (2 casas decimais): usamos format_mb
      sprintf(format_mb, sum(sequence_lengths) / 1000000),
      # Média e Mediana (inteiros): usamos format_int
      sprintf(format_int, round(summary_stats["Mean"])),
      sprintf(format_int, round(summary_stats["Median"])),
      # Desvio Padrão (3 casas decimais): usamos format_float
      sprintf(format_float, stdev),
      # Min, Max, Quartis (inteiros): usamos format_int
      sprintf(format_int, round(summary_stats["Min."])),
      sprintf(format_int, round(summary_stats["Max."])),
      sprintf(format_int, round(quantiles["25%"])),
      sprintf(format_int, round(quantiles["75%"]))
    )
  )

  colnames(summary_df) <- c("Metric", object_label)
  return(summary_df)
}


create_and_plot_frequency_chart <- function(data_subset, column_name, kingdom_label) {
  # Contagem e ordenação dos dados: os dados são ordenados de forma decrescente para que o gráfico destaque as categorias mais frequentes no topo.
  data_summary <- data_subset %>%
    count(.data[[column_name]], name = "Freq") %>%
    arrange(desc(Freq)) %>%
    rename(Entry = 1)

  # Títulos dinâmicos
  main_title <- paste("Diversity of", column_name)
  # subtitle <- paste("Reino (Kingdom):", kingdom_label)

  # Gerar o gráfico
  # A função 'coord_flip()' é utilizada para girar o gráfico, tornando-o horizontal. Isso melhora drasticamente a legibilidade dos longos nomes de Phylums e Families.
  p <- ggplot(data_summary, aes(x = reorder(Entry, Freq), y = Freq, fill = Entry)) +
    geom_bar(stat = "identity") +
    # 'hjust = -0.1' posiciona o rótulo de texto ligeiramente à direita da barra, melhorando a estética do gráfico horizontal.
    geom_text(aes(label = scales::comma(Freq)), hjust = -0.2, size = 4) +
    scale_y_continuous(labels = scales::comma) +
    coord_flip() +
    labs(
      title = main_title,
      # subtitle = subtitle,
      x = column_name,
      y = "Number of Contigs"
    ) +
    theme_minimal(base_size = 12) +
    scale_fill_viridis_d(option = "D") +
    theme(
      legend.position = "none"
    )

  # print(p)
}

PlotPP <- function(ps) {
  # Bind Local Variables
  segstrt <- segend <- segment <- NULL

  # plot variables
  pp.col <- c("red", "gray50", "green", "gray50", "blue")
  vbump <- 0.85

  # motif variables
  # length
  mA <- 11
  mB <- 14
  mC <- 7

  # Segment coordinates
  ## Query
  ps.query <- data.frame(
    segment = c("query"),
    segstrt = 1,
    segend = ps$qlen
  )

  ## Palmprint
  # note this uses a mixture of 1- and 0-base such that
  # motif coordinates are correct 0-base and plotting works
  # in ggplot
  ps.pp <- data.frame(
    segment = c("A", "v1", "B", "v2", "C"),
    segstrt = c(
      ps$pp_start,
      ps$pp_start + mA,
      ps$pp_start + mA + ps$v1_length + 1,
      ps$pp_start + mA + ps$v1_length + mB,
      ps$pp_start + mA + ps$v1_length + mB + ps$v2_length + 1
    ),
    segend = c(
      ps$pp_start + mA,
      ps$pp_start + mA + ps$v1_length + 1,
      ps$pp_start + mA + ps$v1_length + mB,
      ps$pp_start + mA + ps$v1_length + mB + ps$v2_length + 1,
      ps$pp_end
    )
  )

  ppPlot <- ggplot() +
    # Draw Query diagram
    geom_segment(
      data = ps.query,
      aes(x = segstrt, xend = segend, y = 0, yend = 0),
      linetype = 1, size = 4, color = "gray"
    ) +
    xlab("query position (aa)") +
    # Draw Palmprint diagram
    geom_segment(
      data = ps.pp,
      aes(x = segstrt, xend = segend, y = 0, yend = 0, color = segment),
      linetype = 1, size = 10, color = pp.col
    ) +
    geom_text(
      data = ps.pp,
      aes(x = ((segstrt + segend) / 2), y = vbump, label = segment, hjust = "center"),
      color = pp.col
    ) +
    ggtitle(label = paste0(">", ps$query)) +
    ylim(c(-1, 1)) +
    xlim(c(0, NA)) +
    theme(
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      axis.title.y = element_blank(),
      legend.position = "none",
      panel.background = element_blank(),
      panel.border = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      plot.background = element_rect(
        fill = "white",
        colour = "white"
      )
    )

  return(ppPlot)
}
```

```{r logo-setup, echo=FALSE, message=FALSE, warning=FALSE,}
# CRITICAL: This chunk separates the single vector of logo paths (params$logos)
# into two distinct vectors based on the expected filenames defined in the R script.

# Expected pipeline logo filename (must match the R script's EXPECTED_LOGOS)
pipeline_logo_name <- "DeepVir_Logo.png"

# 1. Identify and assign the pipeline logo path
# Corrected: Use the vector of file paths (params$logos) for both filtering and assignment
pipeline_logo_path <- params$logos[grepl(pipeline_logo_name, params$logos, fixed = TRUE)]

# 2. Identify and assign the institutional logo paths
institution_logo_paths <- params$logos[!grepl(pipeline_logo_name, params$logos, fixed = TRUE)]

# Check if the pipeline logo was found
if (length(pipeline_logo_path) == 0) {
  warning("Pipeline logo (DeepVir_Logo.png) not found in the passed paths.")
  # Assign an empty string or placeholder if missing
  pipeline_logo_path <- ""
}
```

```{r setup, message=FALSE, warning=FALSE, echo = FALSE}
# Initial configurations and library loading
library(readr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(scales)
library(viridis)
library(ape)
library(stringr)
library(knitr)

options(digits = 10)

knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# 1. Read FASTA file and clean headers
fasta_expanded_path <- path.expand(params$fasta_file)

if (!file.exists(fasta_expanded_path)) {
  stop(paste("CRITICAL ERROR: Expanded FASTA path does not exist:", fasta_expanded_path))
}

fasta <- read.FASTA(fasta_expanded_path)
names(fasta) <- str_remove_all(names(fasta), pattern = " .+")


output_dir_expanded <- path.expand(params$output_dir)
if (!dir.exists(output_dir_expanded)) {
  dir.create(output_dir_expanded, recursive = TRUE)
}

# CRITICAL FIX: DO NOT MODIFY 'params'. Assign to a new global variable.
OUTPUT_DIR_ABS <- normalizePath(output_dir_expanded, mustWork = TRUE, winslash = "/") # New global variable for file paths

# -----------------------------------------------------------------------------
# 2. Read Taxonomy file (Adaptive: Handles BASTA or Diamond/TaxonKit)
# -----------------------------------------------------------------------------

# Lemos a primeira linha para detectar o formato
first_line <- readLines(params$diamond_file, n = 1)
is_basta <- !grepl("qseqid", first_line) # BASTA não tem header 'qseqid'

if (is_basta) {
  # 1. Ler arquivo BASTA (Sem header, colunas variáveis)
  # Usamos fill=TRUE e stringsAsFactors=FALSE
  raw_data <- read.table(params$diamond_file, sep = "\t", header = FALSE, fill = TRUE, stringsAsFactors = FALSE, quote = "")

  # Renomear para o padrão que o resto do script espera
  colnames(raw_data)[1] <- "qseqid"
  colnames(raw_data)[2] <- "TaxonomyString"

  # 2. Processar a string de taxonomia (separada por ;)
  # Formato esperado: unknown;Phylum;Class;Order;Family;Genus;Species;
  data <- raw_data %>%
    separate(TaxonomyString,
      c("Superkingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"),
      sep = ";",
      fill = "right",
      extra = "drop"
    ) %>%
    mutate(
      # Converter "unknown" ou strings vazias para NA
      across(
        c(Superkingdom, Phylum, Class, Order, Family, Genus, Species),
        ~ na_if(., "")
      ),
      across(
        c(Superkingdom, Phylum, Class, Order, Family, Genus, Species),
        ~ na_if(tolower(.), "unknown")
      ),

      # Criar a coluna 'Overall' para o gráfico de Reinos
      # Tenta pegar Superkingdom -> Phylum -> Class -> Unclassified
      Overall = coalesce(Superkingdom, Phylum, Class, "Unclassified")
    )
} else {
  # Lógica original para arquivos Diamond/TaxonKit com Header
  data <- read.table(params$diamond_file, sep = "\t", fill = T, header = T, na.strings = "", as.is = T)


  data <- data %>%
    mutate(
      Overall = coalesce(as.character(Realm), as.character(Acelular), as.character(Kingdom), "Unclassified")
    )
}


# 3. Split contigs into groups: With Hits (tax) and Without Hits (notax)
# This separation is fundamental for quality analysis. Long contigs without hits (`fasta_notax`) may indicate sequences from novel assemblies or annotation failures, whereas short contigs without hits may be assembly artifacts. Comparing the length metrics between the two groups is critical.
index_fas <- which(names(fasta) %in% data$qseqid)
fasta_tax <- fasta[index_fas]
fasta_notax <- fasta[-index_fas]

# 4. Set up objects for future loops
# 'split_data' stores data subsets by Kingdom, and 'kingdom_names' lists the unique names for iteration in Section 3.
split_data <- split(data, data$Overall)

kingdom_names <- names(split_data)
```

## 🔬 Overview of Contigs Quality and Taxonomic Assignment

This section provides a summary and detailed analysis of contig quality based on successful **taxonomic identification** using the **DIAMOND alignment tool**.

### Quality Metrics and Grouping
Contigs are categorized into three distinct groups for comparative analysis:

1.  **All Contigs:** The complete set of assembled contigs (overall assembly quality).
2.  **Contigs with Hits:** Contigs successfully assigned a taxonomy via DIAMOND.
3.  **Contigs without Hits (No-Hits):** Contigs lacking any successful taxonomic identification.

### Analytical Insight
A **detailed comparison of length metrics** (including **Mean, Median, Minimum, and Maximum** length) across these three groups yields immediate insights into two critical aspects:

* **Assembly Efficacy:** How assembly quality (especially length) correlates with successful database alignment.
* **Data Novelty:** The characteristics of the "No-Hits" group often represent **novel** or highly divergent sequences, which are generally shorter or more fragmented.


```{r contig_quality, message=FALSE, warning=FALSE, echo = FALSE, fig.width=10, fig.height=10, out.width="100%"}
# Length Summary Function
# This function calculates essential metrics like N50 (simulated by Max Len. in simple contexts) and quartiles. Robust initial checking (is.list/length == 0) ensures the script does not halt even if a dataset is empty, returning NAs to maintain the table structure.

# Data Processing and Combination
object_list <- c("fasta", "fasta_tax", "fasta_notax")
summary_results <- list()

for (obj_name in object_list) {
  dnabin_obj <- get(obj_name, envir = .GlobalEnv)
  summary_table_single <- summarize_dnabin_lengths(dnabin_obj, obj_name)
  summary_results[[obj_name]] <- summary_table_single
}

final_summary_table <- summary_results[[1]]
for (i in 2:length(summary_results)) {
  final_summary_table <- merge(final_summary_table, summary_results[[i]], by = "Metric", all = TRUE)
}

colnames(final_summary_table) <- c("Metric", "All Contigs", "Contigs with Hits", "Contigs no Hits")

knitr::kable(
  final_summary_table,
  caption = "Statistics of Assembled Contigs per Taxonomy Category",
  align = c("l", "r", "r", "r")
)

output_filename <- file.path(OUTPUT_DIR_ABS, paste0(params$sample_name, "contigs_summary.tsv"))


write.table(
  final_summary_table,
  file = output_filename,
  sep = "\t",
  row.names = FALSE,
  quote = FALSE # Opcional: evita aspas em torno dos campos de texto
)
```

## 📊 Overall Taxonomic Diversity Analysis
This section analyzes the frequency distribution of all **annotated contigs** at the **broadest taxonomic level** to provide a overall understanding of the sample's composition.

### Scope of Analysis
The analysis is conducted across the four major biological domains:
**Viruses**
**Eukaryota**
**Bacteria**
**Archaea**

### Analytical Goals
Identifying the **dominant** groups and their respective proportions within the sample serves as the critical first step in two main areas:

* **Sample Characterization:** Establishing the baseline composition of the entire dataset.

* **Discovery Assessment: ** Evaluating the overall discovery effort and the distribution of annotated sequences across the major taxonomic domains.

```{r kingdom, message=FALSE, warning=FALSE, echo = FALSE, fig.width=10, fig.height=8, out.width="90%"}
if (nrow(data) > 0) {
  # 1. Count Kingdom frequencies and prepare the table
  data_summary <- as.data.frame(table(data$Overall)) %>%
    rename(Entry = Var1, Freq = Freq) %>%
    mutate(Entry = factor(Entry, levels = Entry[order(Freq, decreasing = TRUE)]))

  # summary table file
  summary_for_table <- data_summary %>%
    mutate(
      Total = sum(Freq),
      Percentage = round((Freq / Total) * 100, 2)
    ) %>%
    arrange(desc(Freq)) %>%
    select(Category = Entry, Count = Freq, Percentage = Percentage)

  knitr::kable(
    summary_for_table,
    caption = "Summary of Contig Hits per Kingdom",
    col.names = c("Category", "Contigs", "Percentage (%)")
  )

  output_filename <- file.path(OUTPUT_DIR_ABS, paste0(params$sample_name, "contigs_summary_per_Kingdom.tsv"))
  # output_filename <- file.path(OUTPUT_DIR_ABS, paste0(params$sample_name,"contigs_summary.tsv"))

  write.table(
    summary_for_table,
    file = output_filename,
    sep = "\t",
    row.names = FALSE,
    quote = FALSE # Opcional: evita aspas em torno dos campos de texto
  )
} else {
  cat("**No contigs with taxonomic classification were found via DIAMOND/Taxonomy.**\n")
}
```
### Bar chart with logarithmic scale. 
It allows for visualizing the vast difference in counts **between Bacteria, Archaea, Eukarya and Viruses**. 
"Virus" counts tend to be much lower, reflecting the nature of their genome - *smaller and less conserved* - and the characteristics of the database used. The simultaneous visualization of these orders of magnitude ensures that minority categories are not visually obscured, although their *low representation requires careful interpretation*.

```{r kingdom_graphs, echo = FALSE, fig.width=10, fig.height=8, out.width="90%"}
if (nrow(data) > 0) {
  # 3. Generate the Global Frequency Plot (KINGDOM plot)
  # The use of 'scale_y_log10' is vital here to visualize the immense count disparity between Superkingdoms. The log scale ensures that low-frequency categories (like viruses) are visible alongside the dominant categories.
  KINGDOM <- ggplot(data_summary, aes(x = Entry, y = Freq, fill = Entry)) +
    geom_bar(stat = "identity") +
    scale_y_log10(labels = scales::comma) +
    geom_text(aes(label = scales::comma(Freq)), vjust = -0.5) +
    labs(
      title = "Principal Classification (Log10)",
      x = "Classification",
      y = "Frequency (Log10)"
    ) +
    theme_minimal(base_size = 14) +
    scale_fill_viridis_d(option = "D") +
    theme(
      legend.position = "none",
      axis.text.x = element_text(angle = 45, hjust = 1)
    )

  print(KINGDOM)

  output_filename <- file.path(OUTPUT_DIR_ABS, paste0(params$sample_name, "KINGDOM_Classification_Log10.png"))

  ggsave(
    filename = output_filename,
    plot = KINGDOM,
    width = 8,
    height = 6,
    dpi = 300
  )
}
```

## 🦠📊 Overall Viral Diversity Analysis

This section analyzes the **frequency distribution** of annotated contigs at the **Viral Family level**.

### Purpose and Importance
* **Characterization:** Identifying the dominant viral groups and their proportions is the critical **first step** in characterizing the sample composition.
* **Discovery Effort:** This analysis provides an overall assessment of the **viral discovery effort** and diversity within the sample.

### Graphical Representation Diversity


```{r contigs-viral-graph, echo=FALSE, fig.width=10, fig.height=15, out.width="90%"}
# 1. Função de Plotagem de Frequência Detalhada (Horizontal Bar Chart)

# AJUSTE DE SEGURANÇA:
# Não tentamos recriar 'Overall' usando Realm/Acelular.
# Apenas refinamos a coluna 'Overall' que já foi criada no setup.
data <- data %>%
  mutate(
    # Garante que NAs virem "Unclassified"
    Overall = replace_na(Overall, "Unclassified"),

    # Padronização: Se a palavra "Viruses" aparecer em qualquer lugar da classificação
    # (ex: "Duplodnaviria (Viruses)"), forçamos a categoria para apenas "Viruses".
    Overall = ifelse(str_detect(Overall, "Viruses"), "Viruses", Overall)
  )

# Filtra apenas o que foi classificado como Vírus
virus_data <- data %>%
  filter(Overall == "Viruses") %>%
  # Se a coluna Family não existir ou for NA, preenche como Unclassified
  mutate(Family = if ("Family" %in% names(.)) replace_na(Family, "Unclassified") else "Unclassified")

# Gera os gráficos e tabelas se houver vírus
if (nrow(virus_data) > 0) {
  output_filename <- file.path(OUTPUT_DIR_ABS, paste0(params$sample_name, "Viral_Diamond.tsv"))
  write.table(virus_data, file = output_filename, sep = "\t", row.names = FALSE, quote = FALSE)

  virus_plot <- create_and_plot_frequency_chart(virus_data, "Family", kingdom_label = "Viruses")
  print(virus_plot)

  output_filename <- file.path(OUTPUT_DIR_ABS, paste0(params$sample_name, "Virus_Classification_Log10.png"))
  ggsave(filename = output_filename, plot = virus_plot, width = 8, height = 6, dpi = 300)
} else {
  cat("**No viral contigs identified in this sample via DIAMOND/Taxonomy.**")
}
```

Contigs determined as **Unclassified** are viruses lacking proper, official taxonomy at the reference level, typically due to: (1)  Non-up-to-date records in public databases (e.g., NCBI); (2) absence of an official classification within the **International Committee on Taxonomy of Viruses (ICTV)**.

### Overall Family Diversity & Richness

```{r contigs_table_viral_family, message=FALSE, echo=FALSE, fig.width=10, fig.height=5, out.width="90%"}
if (nrow(virus_data) > 0) {
  #  get summary of viruses
  sum_virus <- as.data.frame(table(virus_data$Family)) %>%
    rename(Entry = Var1, Freq = Freq)

  # Calculate the grand total count (sum of all frequencies)
  grand_total_count <- sum(sum_virus$Freq)

  # --- 2. Calculate Percentages and Prepare Table ---

  summary_for_table <- sum_virus %>%
    # Recalculate Total for percentage calculation (this is correct here)
    mutate(
      Total_Count = grand_total_count,
      Percentage = round((Freq / Total_Count) * 100, 2)
    ) %>%
    # Arrange by frequency (Count)
    arrange(desc(Freq)) %>%
    # Select and rename columns for final output
    select(Category = Entry, Count = Freq, Percentage)

  # --- 3. Create table of diversity per Viral family id ---

  total_row <- summary_for_table %>%
    # Calculate the sum of Counts and sum of Percentages
    summarise(
      Category = "**TOTAL**",
      Count = sum(Count),
      Percentage = sum(Percentage) # Should be 100.00 or very close
    ) %>%
    # Re-round the total percentage to ensure clean output
    mutate(Percentage = round(Percentage, 2))

  # Combine the summary table with the total row at the end
  final_table <- bind_rows(summary_for_table, total_row)

  # --- 4. Render the Table ---

  knitr::kable(
    final_table,
    caption = "Summary of Contig per Viral Family",
    col.names = c("Family", "Contigs", "Percentage (%)")
  )
} else {
  cat("**No viral data available for Family summary.**")
}
```


## 🦠🧬 Viral Contigs Summary by Family

This table provides an **essential summary** of the assembled contigs, organized by viral family, for **quality assessment (QA)**.

* **Purpose:** To quickly assess the quality of assembled contigs and determine if they represent **near-complete viral genomes** or are candidates of interest.
* **Context:** Different viral families possess **distinct genomic sizes and quality metrics**.
* **Summary Detail:** The table summarizes the **overall assembled contigs** for each identified viral family.

```{r contigs-viral-family, echo=FALSE, fig.width=10, fig.height=15, out.width="90%"}
if (nrow(virus_data) > 0) {
  summary_virus_lengths <- list()
  # Iterate over the UNIQUE names of the families
  for (name in unique(virus_data$Family)) {
    qseqid_for_family <- virus_data$qseqid[which(virus_data$Family %in% name)]
    index <- which(names(fasta) %in% qseqid_for_family)
    in_fasta <- fasta[index]
    summary_table_single <- summarize_dnabin_lengths(in_fasta, name)
    summary_virus_lengths[[name]] <- summary_table_single

    # CRITICAL: Write the per-family FASTA file here as a side effect (optional, but good practice)
    file_name <- file.path(OUTPUT_DIR_ABS, paste0(params$sample_name, "_", name, ".fasta"))
    write.FASTA(in_fasta, file = file_name)
  }

  # Combine all summary tables
  final_length_summary <- summary_virus_lengths[[1]]

  if (length(summary_virus_lengths) > 1) {
    for (i in 2:length(summary_virus_lengths)) {
      final_length_summary <- merge(final_length_summary, summary_virus_lengths[[i]], by = "Metric", all = TRUE)
    }
  }

  output_filename <- file.path(OUTPUT_DIR_ABS, paste0(params$sample_name, "viral_contigs_summary.tsv"))
  write.table(
    final_length_summary,
    file = output_filename,
    sep = "\t",
    row.names = FALSE,
    quote = FALSE
  )

  # Render the Table
  knitr::kable(
    final_length_summary,
    caption = "Length Statistics of Contigs per Viral Family",
    align = c("l", rep("r", length(summary_virus_lengths)))
  )
} else {
  cat("**No viral contigs found to summarize lengths.**")
}
```

## 🦠🌌 Viral Dark Matter Putative Members Iddentification

Among contigs genomic fragments that remain unclassified by standard taxonomic methods, a bioinformatic strategy of iddentification of putative members of the **Viral Dark Matter** was implemented. The primary marker used for identification is the **RNA-dependent RNA Polymerase (RdRp)** sequence. The RdRp gene is highly conserved across numerous RNA viruses, making it an ideal marker for identifying novel or divergent viral sequences characteristic of the VDM.


***Contig Pool**: The starting material consists of non-taxonomically classified contigs derived from metagenomic sequencing.

***ORF Extraction**: Prior to downstream analysis, Open Reading Frames (ORFs) were extracted from the contigs using a comprehensive set of non-standard genetic code tables: 1, 3, 4, 5, 6, 11, and 16. This multi-code approach maximizes the detection of translation products from highly divergent or unusual viral genomes.

***Pipeline**: The strategy to RdRp iddentificatted was thorugh the Palm_annot pipeline ([https://github.com/rcedgar/palm](https://github.com/rcedgar/palm)\_annot.git).


```{r dark_matter_functions, echo=FALSE}
PlotPP <- function(ps) {
  # Bind Local Variables
  segstrt <- segend <- segment <- NULL

  # plot variables
  pp.col <- c("#440154", "gray60", "#3b528b", "gray60", "#fde725")
  vbump <- 0.85

  # --- MODIFICAÇÃO PRINCIPAL AQUI ---
  # O 'ps' é um data frame de 1 linha. Precisamos extrair o vetor da lista.

  # Query
  ps.query <- data.frame(
    segment = c("query"),
    segstrt = 1,
    segend = ps$qlen
  )

  # Palmprint - Extrair os vetores da lista aninhada (usando [[1]])
  ps.pp <- data.frame(
    segment = unlist(ps$order_labels),
    segstrt = unlist(ps$order_starts),
    segend = unlist(ps$order_ends),
    color = unlist(ps$order_colors)
  ) # Criar uma coluna 'color'

  ppPlot <- ggplot() +
    # Draw Query diagram
    geom_segment(
      data = ps.query,
      aes(x = segstrt, xend = segend, y = 0, yend = 0),
      linetype = 1, size = 4, color = "gray"
    ) +
    xlab("query position (aa)") +
    # Draw Palmprint diagram: Use a coluna 'color' que criamos
    geom_segment(
      data = ps.pp,
      aes(x = segstrt, xend = segend, y = 0, yend = 0),
      linetype = 1, size = 10, color = ps.pp$color
    ) + # Usar ps.pp$color
    # Draw labels: Use o dataframe 'ps.pp' com as coordenadas corretas
    geom_text(
      data = ps.pp,
      aes(x = ((segstrt + segend) / 2), y = vbump, label = segment, hjust = "center"),
      color = ps.pp$color
    ) + # Usar ps.pp$color
    ggtitle(label = paste0(">", ps$query)) +
    ylim(c(-1, 1)) +
    xlim(c(0, NA)) +
    theme(
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      axis.title.y = element_blank(),
      legend.position = "none",
      panel.background = element_blank(),
      panel.border = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      plot.background = element_rect(
        fill = "white",
        colour = "white"
      )
    )

  return(ppPlot)
}

# --- PREPARAÇÃO DE DADOS COM LÓGICA DE INVERSÃO ---

# Motifs lengths (hardcoded)
mA <- 11
mB <- 14
mC <- 7

# Cores dos motivos: A, v1, B, v2, C
pp.col.canonical <- c("#440154", "gray60", "#3b528b", "gray60", "#fde725")
```
### Putative RdRp Iddentified

```{r dark_matter_analysis, echo=FALSE,  warning=FALSE, fig.width=10, fig.height=4, out.width="50%"}
# Carregar a tabela
file_info <- file.info(params$darkmatter_file)

if (!is.na(file_info$size) && file_info$size > 0) {
  # Carrega a tabela com tryCatch para evitar falhas caso o arquivo contenha apenas linhas em branco
  fev <- tryCatch(
    {
      read.table(file = params$darkmatter_file, header = TRUE, sep = "\t")
    },
    error = function(e) {
      data.frame() # Retorna dataframe vazio em caso de erro de leitura
    }
  )
} else {
  fev <- data.frame() # Arquivo vazio gera dataframe vazio
}

# AQUI ESTA A CORRECAO: Verificar se a tabela tem dados
if (nrow(fev) > 0) {
  # 1. Mapeamento de Colunas Fixas
  fev$query <- fev$Label
  fev$qlen <- nchar(fev$aaseq)

  # 2. Lógica Condicional para Ordem (ABC vs CAB)
  fev$order_labels <- list(NA)
  fev$order_colors <- list(NA)
  fev$order_starts <- list(NA)
  fev$order_ends <- list(NA)

  # Função para calcular as coordenadas e comprimentos baseados na ordem
  for (i in 1:nrow(fev)) {
    ps <- fev[i, ]

    # 2a. Definir os pontos de início dos motivos centrais
    if (ps$pssm_ABC == "ABC") {
      # Ordem Canônica: A -> B -> C
      pA <- ps$posA
      pB <- ps$posB
      pC <- ps$posC

      # Calcular os comprimentos das regiões variáveis
      v1_len <- pB - (pA + mA)
      v2_len <- pC - (pB + mB)

      # Sequência de Motivos na Ordem de Plotagem
      labels <- c("A", "v1", "B", "v2", "C")
      colors <- pp.col.canonical

      # Coordenadas (baseadas em A -> B -> C)
      starts <- c(pA, pA + mA + 1, pB, pB + mB + 1, pC)
      ends <- c(pA + mA, pB, pB + mB, pC, pC + mC - 1)
    } else if (ps$pssm_ABC == "CAB") {
      # Ordem Invertida: C -> A -> B
      pC <- ps$posC
      pA <- ps$posA
      pB <- ps$posB

      v1_len <- pA - (pC + mC)
      v2_len <- pB - (pA + mA)

      labels <- c("C", "v1", "A", "v2", "B")
      colors <- c(pp.col.canonical[5], pp.col.canonical[2], pp.col.canonical[1], pp.col.canonical[4], pp.col.canonical[3])

      starts <- c(pC, pC + mC + 1, pA, pA + mA + 1, pB)
      ends <- c(pC + mC, pA, pA + mA, pB, pB + mB)
    } else {
      next
    }

    all_starts <- c(pA, pB, pC)
    all_ends <- c(pA + mA, pB + mB, pC + mC)

    ps$pp_start <- min(all_starts)
    ps$pp_end <- max(all_ends)
    ps$v1_length <- v1_len
    ps$v2_length <- v2_len

    fev[i, ]$pp_start <- ps$pp_start
    fev[i, ]$pp_end <- ps$pp_end
    fev[i, ]$v1_length <- ps$v1_length
    fev[i, ]$v2_length <- ps$v2_length

    fev[i, ]$order_labels <- list(labels)
    fev[i, ]$order_colors <- list(colors)
    fev[i, ]$order_starts <- list(starts)
    fev[i, ]$order_ends <- list(ends)
  }


  # --- EXECUÇÃO DO PLOT ---
  for (i in 1:nrow(fev)) {
    pp.plot <- PlotPP(fev[i, , drop = FALSE])

    output_filename <- file.path(OUTPUT_DIR_ABS, paste0(params$sample_name, "_", fev$query[i], "RdRp_Motifs.png"))

    ggsave(
      filename = output_filename,
      plot = pp.plot,
      width = 8,
      height = 6,
      dpi = 300
    )

    print(pp.plot)
  }
} else {
  cat("**No putative RdRp sequences identified in this sample (Dark Matter Analysis).**")
}
```


## Considerations
To ensure the high confidence and biological relevance of the identified VDM contigs, a dual-strategy validation approach must be further implemented to characterize their origin and potential function.

#### 1. Host Transcript 

Validation within the **Blastn* (https://blast.ncbi.nlm.nih.gov/Blast.cgi) to comprehensively screen the putative VDM contig sequences against the vast database of Genbank hostranscripts. This step is critical for **robustly confirming** the viral provenance of the VDM contigs, specifically by identifying missassembled contigs as VDM whereas its true nature may be a non translatable host transcript, absent in blastx approachs.

* **Tool:** Blastn (https://blast.ncbi.nlm.nih.gov/Blast.cgi)
* **Purpose:** Confirmation of **viral origin** within the contig.
* **Expectations:** No hits with hosts transcripts within Blastn clustered database.


#### 2. Protein Model Filtering (Phyre2)

To mitigate the risk of **false positives** arising from misidentified **Host Protein Models**, subsequent analysis employed the **Phyre2** server. Phyre2 was used to perform **homology modeling and remote structural homology detection** on the translations of the VDM contigs. Any contigs showing high confidence structural similarity (e.g., >90% confidence score) to known cellular/host proteins must be **disconsered**.

* **Tool:** Phyre2 (https://www.sbg.bio.ic.ac.uk/phyre2)
* **Purpose:** Exclusion of **Host Proteins**.


### Pipeline

```{r incluir-logos-DeepVir, echo=FALSE, fig.show='hold', out.width='50%', out.height='10%'}
# Use the pipeline_logo_path variable defined in the logo-setup chunk
if (length(pipeline_logo_path) > 0 && pipeline_logo_path != "") {
  knitr::include_graphics(pipeline_logo_path)
}
```

### Institutions

```{r incluir-logos-horizontal, echo=FALSE, fig.show='hold', out.width='20%', out.height='20%'}
# Este chunk insere as 4 imagens e controla o tamanho e o layout.
# 'echo=FALSE' esconde o código R no relatório final.
# 'fig.show='hold'' é crucial para tentar manter todas as imagens na mesma linha.
# 'out.width='25%'' define que cada imagem terá 25% da largura da página.
# 'fig.align='center'' centraliza o grupo de imagens.
# Access the list of paths passed via the parameter
# Use the institution_logo_paths variable defined in the logo-setup chunk
if (length(institution_logo_paths) > 0) {
  # knitr::include_graphics automatically handles a vector of paths
  knitr::include_graphics(institution_logo_paths)
}
```
